---
title: "Sociodemographic Analysis"
author: "Natália Wirowski"
date: "2026-08-31"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
---
## Importing metadata

```{r}
library(readr)
metadata <- read_delim("~/data/metadata.txt", 
    delim = "\t", escape_double = FALSE, 
    trim_ws = TRUE)
```

Qualitative variables to be used:
Variable name - Label name - Levels
thdico - Group - 0: "Control", 1: "Mood Disorder"
corpele - Skin Color - 1: "White", 2: "Non-White"
sexo - Sex - 1: "Male", 2: "Female"
ABEP - Socioeconomic status - 1: "Upper class", 2: "Lower class"
histpsifam - History of familial psychiatric disorder - 1: "Yes", 0: "No"
a20medicpsi - Use of psychiatric medication - 1: "Yes", 0: "No"
antib - Use of antibiotics in the last monthn - 1: "Yes", 0: "No"
suplem - Use of prebiotics, probiotics, or synbiotics in the last month - 1: "Yes", 0: "No"

Quantitative variables to be used:
idade - Age
anosestudo - Years of schooling
IMC - Body Mass Index

We need to make sure the categorical variables are labeled correctly, plus levels.

## Labeling
```{r}
metadata$thdico <- factor(metadata$thdico,
                        levels = c(0, 1),
                        labels = c("Control", "Mood Disorder"))

metadata$corpele <- factor(metadata$corpele,
                           levels = c(1, 2),
                           labels = c("White", "Non-White"))

metadata$sexo <- factor(metadata$sexo,
                        levels = c(1, 2),
                        labels = c("Male", "Female"))

metadata$ABEP <- factor(metadata$ABEP,
                        levels = c(1, 2),
                        labels = c("Upper class", "Lower class"))

metadata$histpsifam <- factor(metadata$histpsifam,
                              levels = c(1, 0), 
                              labels = c("Yes", "No"))

metadata$a20medicpsi <- factor(metadata$a20medicpsi,
                              levels = c(1, 0), 
                              labels = c("Yes", "No"))

metadata$antib <- factor(metadata$antib,
                              levels = c(1, 0), 
                              labels = c("Yes", "No"))

metadata$suplem <- factor(metadata$suplem,
                              levels = c(1, 0), 
                              labels = c("Yes", "No"))

metadata$idade <- as.numeric(metadata$idade)
metadata$anosestudo <- as.numeric(metadata$anosestudo)
metadata$IMC <- as.numeric(metadata$IMC)

```

## Assessing the normality of quantitative variables

Descriptive statistics - Age

```{r}
mean(metadata$idade)  
median(metadata$idade)
sd(metadata$idade)
```

Descriptive statistics - Years of schooling

```{r}
mean(metadata$anosestudo)  
median(metadata$anosestudo) 
sd(metadata$anosestudo) 
```

Descriptive statistics - BMI

```{r}
mean(metadata$IMC)  
median(metadata$IMC) 
sd(metadata$IMC) 
```

Histogram - Age

```{r}
hist(metadata$idade)
```


```{r}
breaks_seq <- seq(min(metadata$idade, na.rm = TRUE) - 0.5, 
                  max(metadata$idade, na.rm = TRUE) + 0.5, 
                  by = 1)

hist(metadata$idade,
     breaks = breaks_seq,
     probability = TRUE,  
     main = "Histogram of Age with Gaussian Curve",
     xlab = "Age",
     col = "skyblue",
     border = "white",
     xlim = c(25,40)) 

# Adicionar curva normal
mean_idade <- mean(metadata$idade, na.rm = TRUE)
sd_idade <- sd(metadata$idade, na.rm = TRUE)
curve(dnorm(x, mean = mean_idade, sd = sd_idade), 
      col = "red", 
      lwd = 2, 
      add = TRUE)

```
Histogram - Years of schooling

```{r}
hist(metadata$anosestudo)
```

```{r}
breaks_seq <- seq(min(metadata$anosestudo, na.rm = TRUE) - 0.5, 
                  max(metadata$anosestudo, na.rm = TRUE) + 0.5, 
                  by = 1)

hist(metadata$anosestudo,
     breaks = breaks_seq,
     probability = TRUE,  
     main = "Histogram of Years of Schooling",
     xlab = "Years of Schooling",
     col = "lightgreen",
     border = "white",
     xlim = c(0, 20))  # ajuste o eixo x se desejar

mean_anos <- mean(metadata$anosestudo, na.rm = TRUE)
sd_anos <- sd(metadata$anosestudo, na.rm = TRUE)
curve(dnorm(x, mean = mean_anos, sd = sd_anos), 
      col = "red", 
      lwd = 2, 
      add = TRUE)

```

Histogram - BMI


```{r}
hist(metadata$IMC)
```

```{r}
range(metadata$IMC, na.rm = TRUE)
```


```{r}
breaks_imc <- seq(
  floor(min(metadata$IMC, na.rm = TRUE)) - 0.5,
  ceiling(max(metadata$IMC, na.rm = TRUE)) + 0.5,
  by = 1
)

hist(
  metadata$IMC,
  breaks = breaks_imc,
  probability = TRUE,
  main = "Histogram of BMI with Gaussian Curve",
  xlab = "Body Mass Index",
  col = "lightgreen",
  border = "white"
)

mean_imc <- mean(metadata$IMC, na.rm = TRUE)
sd_imc <- sd(metadata$IMC, na.rm = TRUE)

curve(
  dnorm(x, mean = mean_imc, sd = sd_imc),
  col = "red",
  lwd = 2,
  add = TRUE
)

```



## Parametric tests

Mean and standard deviation
```{r}
library(tidyverse)
mean_sd_table <- metadata %>%
  group_by(thdico) %>%
  summarise(
    mean_age = mean(idade, na.rm = TRUE),
    sd_age = sd(idade, na.rm = TRUE),
    mean_school = mean(anosestudo, na.rm = TRUE),
    sd_school = sd(anosestudo, na.rm = TRUE),
    mean_BMI = mean(IMC, na.rm = TRUE),
    sd_BMI = sd(IMC, na.rm = TRUE)
  )

mean_sd_table
```

Total

```{r}
mean_sd_table2 <- metadata %>%
  summarise(
    mean_age = mean(idade, na.rm = TRUE),
    sd_age = sd(idade, na.rm = TRUE),
    mean_school = mean(anosestudo, na.rm = TRUE),
    sd_school = sd(anosestudo, na.rm = TRUE),
    mean_BMI = mean(IMC, na.rm = TRUE),
    sd_BMI = sd(IMC, na.rm = TRUE)
  )

mean_sd_table2
```


## t test
```{r}
t.test(idade ~ thdico, data = metadata)

t.test(anosestudo ~ thdico, data = metadata)

t.test(IMC ~ thdico, data = metadata)
```
## Non Parametric tests

Median and interquartile range

```{r}
median_iqr_table <- metadata %>%
  group_by(thdico) %>%
  summarise(
    median_age = median(idade, na.rm = TRUE),
    IQR_age = sprintf(
      "%.2f–%.2f",
      quantile(idade, 0.25, na.rm = TRUE),
      quantile(idade, 0.75, na.rm = TRUE)
    ),
    median_school = median(anosestudo, na.rm = TRUE),
    IQR_school = sprintf(
      "%.2f–%.2f",
      quantile(anosestudo, 0.25, na.rm = TRUE),
      quantile(anosestudo, 0.75, na.rm = TRUE)
    ),
    median_IMC = median(IMC, na.rm = TRUE),
    IQR_IMC = sprintf(
      "%.2f–%.2f",
      quantile(IMC, 0.25, na.rm = TRUE),
      quantile(IMC, 0.75, na.rm = TRUE)
    )
  )

median_iqr_table
```

Total

```{r}
median_iqr_table2 <- metadata %>%
    summarise(
      median_age = median(idade, na.rm = TRUE),
      IQR_age = sprintf(
        "%.2f–%.2f",
        quantile(idade, 0.25, na.rm = TRUE),
        quantile(idade, 0.75, na.rm = TRUE)
      ),
      median_school = median(anosestudo, na.rm = TRUE),
      IQR_school = sprintf(
        "%.2f–%.2f",
        quantile(anosestudo, 0.25, na.rm = TRUE),
        quantile(anosestudo, 0.75, na.rm = TRUE)
      ),
      median_IMC = median(IMC, na.rm = TRUE),
      IQR_IMC = sprintf(
        "%.2f–%.2f",
        quantile(IMC, 0.25, na.rm = TRUE),
        quantile(IMC, 0.75, na.rm = TRUE)
      )
    )

median_iqr_table2
```


## Mann-Whitney (Wilcoxon rank-sum test)
```{r}
wilcox.test(idade ~ thdico, data = metadata)

wilcox.test(anosestudo ~ thdico, data = metadata)

wilcox.test(IMC ~ thdico, data = metadata)
```

```{r}
wilcox.test(idade ~ thdico, data = metadata, correct = FALSE)

wilcox.test(anosestudo ~ thdico, data = metadata, correct = FALSE)

wilcox.test(IMC ~ thdico, data = metadata, correct = FALSE)
```


## Categorical variables table

Absolute and relative frequency table

```{r}
library(janitor)

freq_table <- function(data, var) {
  
  data %>%
    filter(!is.na({{ var }})) %>%
    tabyl({{ var }}, thdico) %>%
    adorn_totals(where = c("row", "col")) %>%
    adorn_percentages(denominator = "col") %>%
    adorn_pct_formatting(digits = 1) %>%
    adorn_ns(position = "front") %>%
    select({{ var }}, Total, `Mood Disorder`, Control)
}

tab_corpele <- freq_table(metadata, corpele)
tab_sexo <- freq_table(metadata, sexo)
tab_ABEP <- freq_table(metadata, ABEP)
tab_histpsifam <- freq_table(metadata, histpsifam)

tab_a20medicpsi <- freq_table(metadata, a20medicpsi)
tab_antib <- freq_table(metadata, antib)
tab_suplem <- freq_table(metadata, suplem)

tab_corpele
tab_sexo
tab_ABEP
tab_histpsifam

tab_a20medicpsi
tab_antib
tab_suplem

```
## Chi-square

Sex by thdico
```{r}
tab_sexo <- table(metadata$sexo, metadata$thdico)
tab_sexo
```

```{r}
chisq.test(tab_sexo)$expected
```


```{r}
chisq.test(tab_sexo)
chisq.test(tab_sexo, correct = FALSE)
fisher.test(tab_sexo)
```
Skin color by thdico

```{r}
tab_corpele <- table(metadata$corpele, metadata$thdico)
tab_corpele
```

```{r}
chisq.test(tab_corpele)$expected
```

```{r}
chisq.test(tab_corpele)
chisq.test(tab_corpele, correct = FALSE)
fisher.test(tab_corpele)
```

Socioeconomic status by thdico

```{r}
tab_abep <- table(metadata$ABEP, metadata$thdico)
tab_abep
```

```{r}
chisq.test(tab_abep)$expected
```

```{r}
chisq.test(tab_abep)
chisq.test(tab_abep, correct = FALSE)
fisher.test(tab_abep)
```

Family history by thdico

```{r}
tab_hist <- table(metadata$histpsifam, metadata$thdico)
tab_hist
```

```{r}
chisq.test(tab_hist)$expected
```

```{r}
chisq.test(tab_hist)
chisq.test(tab_hist, correct = FALSE)
fisher.test(tab_hist)
```

Psychiatric medication

```{r}
tab_med <- table(
  metadata$a20medicpsi,
  metadata$thdico
)

tab_med
```

```{r}
chisq.test(tab_med)$expected
```

```{r}
chisq.test(tab_med)
chisq.test(tab_med, correct = FALSE)
fisher.test(tab_med)
```


Antibiotics

```{r}
tab_antib <- table(
  metadata$antib,
  metadata$thdico
)

tab_antib
```

```{r}
chisq.test(tab_antib)$expected
```


```{r}
chisq.test(tab_antib)
chisq.test(tab_antib, correct = FALSE)
fisher.test(tab_antib)
```


Prebiotic/probiotic/synbiotic

```{r}
tab_suplem <- table(
  metadata$suplem,
  metadata$thdico
)

tab_suplem
```

```{r}
chisq.test(tab_suplem)$expected
```


```{r}
chisq.test(tab_suplem)
chisq.test(tab_suplem, correct = FALSE)
fisher.test(tab_suplem)
```

