---
title: "Sociodemographic Analysis"
author: "Natália Wirowski"
date: "2025-10-16"
output:
  word_document: default
  html_document:
    df_print: paged
---
## Importing metadata

```{r}
library(readr)
metadata <- read_delim("~/data/metadata.txt", 
    delim = "\t", escape_double = FALSE, 
    trim_ws = TRUE)
```

Qualitative variables to be used:
Variable name - Label name - Levels
mood - Group - 0: "Control", 2: "Bipolar", 3: "Depression"
corpele - Skin Color - 1: "White", 2: "Non-White"
sexo - Sex - 1: "Male", 2: "Female"
ABEP - Socioeconomic status - 1: "Upper class", 2: "Lower class"
histpsifam - History of familial psychiatric disorder - 1: "Yes", 0: "No"

Quantitative variables to be used:
idade - Age
anosestudo - Years of schooling

We need to make sure the categorical variables are labeled correctly, plus levels.

## Labeling
```{r}
metadata$mood <- factor(metadata$mood,
                        levels = c(0, 2, 3),
                        labels = c("Control", "Bipolar", "Depression"))

metadata$corpele <- factor(metadata$corpele,
                           levels = c(1, 2),
                           labels = c("White", "Non-White"))

metadata$sexo <- factor(metadata$sexo,
                        levels = c(1, 2),
                        labels = c("Male", "Female"))

metadata$ABEP <- factor(metadata$ABEP,
                        levels = c(1, 2),
                        labels = c("Upper class", "Lower class"))

metadata$histpsifam <- factor(metadata$histpsifam,
                              levels = c(1, 0), 
                              labels = c("Yes", "No"))

metadata$idade <- as.numeric(metadata$idade)
metadata$anosestudo <- as.numeric(metadata$anosestudo)

```

## Assessing the normality of quantitative variables

Descriptive statistics - Age

```{r}
mean(metadata$idade)  
median(metadata$idade)
sd(metadata$idade)
```

Descriptive statistics - Years of schooling

```{r}
mean(metadata$anosestudo)  
median(metadata$anosestudo) 
sd(metadata$anosestudo) 
```
Histogram - Age

```{r}
hist(metadata$idade)
```


```{r}
breaks_seq <- seq(min(metadata$idade, na.rm = TRUE) - 0.5, 
                  max(metadata$idade, na.rm = TRUE) + 0.5, 
                  by = 1)

hist(metadata$idade,
     breaks = breaks_seq,
     probability = TRUE,  
     main = "Histogram of Age with Gaussian Curve",
     xlab = "Age",
     col = "skyblue",
     border = "white",
     xlim = c(25,40)) 

# Adicionar curva normal
mean_idade <- mean(metadata$idade, na.rm = TRUE)
sd_idade <- sd(metadata$idade, na.rm = TRUE)
curve(dnorm(x, mean = mean_idade, sd = sd_idade), 
      col = "red", 
      lwd = 2, 
      add = TRUE)

```
Histogram - Years of schooling

```{r}
hist(metadata$anosestudo)
```

```{r}
breaks_seq <- seq(min(metadata$anosestudo, na.rm = TRUE) - 0.5, 
                  max(metadata$anosestudo, na.rm = TRUE) + 0.5, 
                  by = 1)

hist(metadata$anosestudo,
     breaks = breaks_seq,
     probability = TRUE,  
     main = "Histogram of Years of Schooling",
     xlab = "Years of Schooling",
     col = "lightgreen",
     border = "white",
     xlim = c(0, 20))  # ajuste o eixo x se desejar

mean_anos <- mean(metadata$anosestudo, na.rm = TRUE)
sd_anos <- sd(metadata$anosestudo, na.rm = TRUE)
curve(dnorm(x, mean = mean_anos, sd = sd_anos), 
      col = "red", 
      lwd = 2, 
      add = TRUE)

```
## Parametric tests

Mean and standard deviation
```{r}
library(tidyverse)
mean_sd_table <- metadata %>%
  group_by(mood) %>%
  summarise(
    mean_age = mean(idade),
    sd_age = sd(idade),
    mean_school = mean(anosestudo),
    sd_school = sd(anosestudo)
  )

mean_sd_table
```

Total

```{r}
library(tidyverse)
mean_sd_table2 <- metadata %>%
  summarise(
    mean_age = mean(idade),
    sd_age = sd(idade),
    mean_school = mean(anosestudo),
    sd_school = sd(anosestudo)
  )

mean_sd_table2
```


## ANOVA
```{r}
anova_idade <- aov(idade ~ mood, data = metadata)
summary(anova_idade)
```

```{r}
anova_anos <- aov(anosestudo ~ mood, data = metadata)
summary(anova_anos)
```
## Non Parametric tests

Median and interquartile range

```{r}
median_iqr_table <- metadata %>%
  group_by(mood) %>%
  summarise(
    median_age = median(idade, na.rm = TRUE),
    IQR_age = sprintf("%.2f–%.2f",
                      quantile(idade, 0.25, na.rm = TRUE),
                      quantile(idade, 0.75, na.rm = TRUE)),
    median_school = median(anosestudo, na.rm = TRUE),
    IQR_school = sprintf("%.2f–%.2f",
                         quantile(anosestudo, 0.25, na.rm = TRUE),
                         quantile(anosestudo, 0.75, na.rm = TRUE))
  )

median_iqr_table
```

Total

```{r}
median_iqr_table2 <- metadata %>%
  summarise(
    median_age = median(idade, na.rm = TRUE),
    IQR_age = sprintf("%.2f–%.2f",
                      quantile(idade, 0.25, na.rm = TRUE),
                      quantile(idade, 0.75, na.rm = TRUE)),
    median_school = median(anosestudo, na.rm = TRUE),
    IQR_school = sprintf("%.2f–%.2f",
                         quantile(anosestudo, 0.25, na.rm = TRUE),
                         quantile(anosestudo, 0.75, na.rm = TRUE))
  )

median_iqr_table2
```


## Kruskal-Wallis
```{r}
kruskal_idade <- kruskal.test(idade ~ mood, data = metadata)
kruskal_idade
```

```{r}
kruskal_anos <- kruskal.test(anosestudo ~ mood, data = metadata)
kruskal_anos
```
## Categorical variables table

Absolute and relative frequency table

```{r}
library(janitor)

freq_table <- function(data, var) {
  data %>%
    tabyl({{ var }}, mood) %>%            
    adorn_totals(where = c("row", "col")) %>%  
    adorn_percentages(denominator = "col") %>% 
    adorn_pct_formatting(digits = 1) %>%     
    adorn_ns()                                 
}


tab_corpele   <- freq_table(metadata, corpele)
tab_sexo      <- freq_table(metadata, sexo)
tab_ABEP      <- freq_table(metadata, ABEP)
tab_histpsifam <- freq_table(metadata, histpsifam)


tab_corpele
tab_sexo
tab_ABEP
tab_histpsifam

```
## Chi-square

Age by mood
```{r}
tab_sexo <- table(metadata$sexo, metadata$mood)
tab_sexo
```

```{r}
chisq.test(tab_sexo)$expected
```


```{r}
chisq.test(tab_sexo)
fisher.test(tab_sexo)
```
Skin color by mood

```{r}
tab_corpele <- table(metadata$corpele, metadata$mood)
tab_corpele
```

```{r}
chisq.test(tab_corpele)$expected
```

```{r}
chisq.test(tab_corpele)
fisher.test(tab_corpele)
```

Socioeconomic status by mood

```{r}
tab_abep <- table(metadata$ABEP, metadata$mood)
tab_abep
```

```{r}
chisq.test(tab_abep)$expected
```


```{r}
chisq.test(tab_abep)
fisher.test(tab_abep)
```
History of familial psychiatric disorder by mood

```{r}
tab_hist <- table(metadata$histpsifam, metadata$mood)
tab_hist
```

```{r}
chisq.test(tab_hist)$expected
```

```{r}
chisq.test(tab_hist)
fisher.test(tab_hist)
```

